---
title: "AquaSensR inputs and checks"
vignette: >
  %\VignetteIndexEntry{inputs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(AquaSensR)
```

AquaSensR requires two input files to use the functions in the package:

1. **Continuous monitoring data** — time series of sensor observations, one column per parameter.
2. **QC threshold metadata** — site- and parameter-specific thresholds used by the four QC checks (gross range, spike, rate of change, and flatline).

Both file types are Excel workbooks (`.xlsx`). This vignette describes how to import and validate each input dataset using the `readASR*` and `checkASR*` family of functions. Example files are included with the package and are used throughout.

## Load the package

Load the package in an R session after installation: 

```{r}
#| eval: false
library(AquaSensR)
```

## File paths

In practice you will supply paths to your own files, for example:

```{r}
#| eval: false
contpth <- "path/to/your/ContinuousData.xlsx"
metapth <- "path/to/your/Metadata.xlsx"
```

The examples below use the files included with the package:

```{r}
contpth <- system.file("extdata/ExampleCont.xlsx", package = "AquaSensR")
metapth <- system.file("extdata/ExampleMeta.xlsx", package = "AquaSensR")
```

---

## Continuous monitoring data

Use `readASRcont()` to import continuous monitoring data. The function reads the Excel file, runs a series of checks via `checkASRcont()`, and then formats the result for downstream use. The `tz` argument sets the time zone for the combined `DateTime` column (see `OlsonNames()` for valid values).

```{r}
contdat <- readASRcont(contpth, tz = "Etc/GMT+5")
```

### Format requirements

The workbook must contain the following columns (additional unrecognised columns will trigger an error):

| Column | Description |
|--------|-------------|
| `Site` | Site identifier |
| `Date` | Observation date, parseable by `lubridate::ymd()` (e.g., `2024-06-01`) |
| `Time` | Observation time as a full datetime string (e.g., `HH:MM:SS`), parseable by `lubridate::ymd_hms()` |
| At least one parameter column | Column name must match a `Parameter` entry in `paramsASR` (e.g., `Water Temp_C`) |

### Checks performed

`checkASRcont()` verifies the following and stops with an informative error if any check fails:

1. **Column names** — all columns are either `Site`, `Date`, `Time`, or a recognised parameter from `paramsASR`.
2. **Required columns present** — `Site`, `Date`, and `Time` exist.
3. **At least one parameter column** — at least one column matches an entry in `paramsASR$Parameter`.
4. **Date format** — all values in `Date` parse successfully with `lubridate::ymd()`.
5. **Time format** — all values in `Time` parse successfully with `lubridate::ymd_hms()`.
6. **No missing values** — no `NA` in any column.
7. **Numeric parameter columns** — all parameter columns contain numeric values.

### Example: triggering an error

Adding an unrecognised column causes `checkASRcont()` to stop immediately:

```{r}
#| error: true
library(dplyr)

contdat_raw <- suppressWarnings(
  readxl::read_excel(contpth, na = c("NA", "na", ""), guess_max = Inf)
) |>
  mutate(across(
    where(~ inherits(.x, "POSIXct") | inherits(.x, "Date")),
    as.character
  ))

# add a column that is not in paramsASR
contdat_raw$BadColumn <- 1

checkASRcont(contdat_raw)
```

### Output format

After passing all checks, `readASRcont()` returns a data frame with:

- `Site` — site identifier (character)
- `DateTime` — combined and time-zone-aware `POSIXct` column
- One numeric column per parameter present in the input file

```{r}
head(contdat)
```

---

## QC threshold metadata

Use `readASRmeta()` to import the QC threshold metadata. The function reads the workbook, runs checks via `checkASRmeta()`, and returns a formatted data frame.

```{r}
metadat <- readASRmeta(metapth)
```

### Format requirements

The workbook must contain exactly the following columns (all required; thresholds you do not want to apply should be left blank / `NA`):

| Column | Description |
|--------|-------------|
| `Site` | Site identifier matching those in the continuous data |
| `Parameter` | Parameter name matching `paramsASR$Parameter` |
| `Depth` | Sensor deployment depth (numeric; can be `NA`) |
| `GrMinFail` | Gross range — lower fail threshold |
| `GrMaxFail` | Gross range — upper fail threshold |
| `GrMinSuspect` | Gross range — lower suspect threshold |
| `GrMaxSuspect` | Gross range — upper suspect threshold |
| `SpikeFail` | Spike — absolute step size for a fail flag |
| `SpikeSuspect` | Spike — absolute step size for a suspect flag |
| `FlatFailN` | Flatline — consecutive identical observations for a fail flag |
| `FlatFailDelta` | Flatline — maximum within-run absolute change treated as "identical" for fail |
| `FlatSuspectN` | Flatline — consecutive identical observations for a suspect flag |
| `FlatSuspectDelta` | Flatline — maximum within-run absolute change treated as "identical" for suspect |
| `RoCN` | Rate of change — multiplier applied to the rolling SD (flag if `\|diff\| > SD × RoCN`) |
| `RoCHours` | Rate of change — look-back window length in hours |

### Checks performed

`checkASRmeta()` verifies the following and stops with an informative error if any check fails:

1. **Column names** — all columns are in the required list above.
2. **All columns present** — every required column exists.
3. **At least one parameter** — at least one value in `Parameter` matches `paramsASR$Parameter`.
4. **Parameter format** — all `Parameter` values match those in `paramsASR$Parameter`.
5. **Numeric columns** — all columns except `Site` and `Parameter` contain only numeric or missing values.

### Example: triggering an error

Supplying an unrecognised parameter name fails the parameter format check:

```{r}
#| error: true
metadat_raw <- suppressWarnings(
  readxl::read_excel(metapth, na = c("NA", "na", ""), guess_max = Inf)
)

# introduce a typo in the Parameter column
metadat_raw$Parameter[1] <- "WaterTemp"

checkASRmeta(metadat_raw)
```

### Output format

After passing all checks, `readASRmeta()` returns a data frame with the columns listed in the format requirements table above, with all threshold columns coerced to numeric.

```{r}
head(metadat)
```