---
title: "AquaSensR inputs and checks"
vignette: >
  %\VignetteIndexEntry{inputs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
library(AquaSensR)
library(dplyr)
```

AquaSensR requires two input files to use the functions in the package:

1. **Continuous monitoring data**: time series of sensor observations, one column per parameter.
2. **QC threshold metadata**: site- and parameter-specific thresholds used by the four QC checks (gross range, spike, rate of change, and flatline).

Both file types are Excel workbooks (`.xlsx`). This vignette describes how to import and check each input dataset. It is critical that the input datasets follow the exact specified format.  Example files with the correct format are included with the package and are used throughout.

## Load the package

Load the package in an R session after installation: 

```{r}
#| eval: false
library(AquaSensR)
```

## File paths

First, specify the location of the two files by saving their paths to R variables. In practice you will supply paths to your own files, for example:

```{r}
#| eval: false
contpth <- "path/to/your/ContinuousData.xlsx"
metapth <- "path/to/your/Metadata.xlsx"
```

The examples below use the files included with the package:

```{r}
contpth <- system.file("extdata/ExampleCont.xlsx", package = "AquaSensR")
metapth <- system.file("extdata/ExampleMeta.xlsx", package = "AquaSensR")
```

## Continuous monitoring data

Use `readASRcont()` to import continuous monitoring data. The function reads the Excel file, automatically runs a series of checks via `checkASRcont()`, and then formats the result for downstream use. The `tz` argument sets the time zone for the combined `DateTime` column (see `OlsonNames()` for valid values).  Be careful to specify the correct time zone, particularly one using a fixed offset (e.g., `Etc/GMT+5` for Eastern time) to avoid issues with daylight saving time transitions. If your data are in local time and the time zone observes DST, consider using a time zone like `America/New_York` that will automatically adjust for daylight savings.

```{r}
contdat <- readASRcont(contpth, tz = "Etc/GMT+5")
```

### Format requirements

The continuous monitoring data workbook must contain the following columns (additional unrecognised columns will trigger an error):

| Column | Description |
|--------|-------------|
| `Site` | Site identifier |
| `Date` | Observation date, parseable by `lubridate::ymd()` (e.g., `2024-06-01`) |
| `Time` | Observation time as a full datetime string (e.g., `HH:MM:SS`), parseable by `lubridate::ymd_hms()` |
| At least one parameter column | Column name must match a `Parameter` entry in `paramsASR` (e.g., `Water Temp_C`) |

Currently, AquaSensR allows the following parameters.  Note the inclusion of the units in the parameter name.  Make sure the parameter name matches the units used in your data.

```{r, echo = F}
paramsASR |>
  select(
    Description = `Label`,
    `Required file name` = `Parameter`,
    Units = `uom`
  ) |>
  knitr::kable()
```

The list above can also be viewed in R with the `paramsASR` dataset, which is included in the package and used for the checks. 

```{r}
paramsASR
```

### Checks performed

The `readASRcont()` function imports the data and runs a series of checks using the `checkASRcont()` function.  The checks evaluate the following and stops with an informative error if any check fails:

1. **Column names**: all columns are either `Site`, `Date`, `Time`, or a recognised parameter from `paramsASR`.
2. **Required columns present**: `Site`, `Date`, and `Time` exist.
3. **At least one parameter column**: at least one column matches an entry in `paramsASR$Parameter`.
4. **Date format**: all values in `Date` parse successfully with `lubridate::ymd()`.
5. **Time format**: all values in `Time` parse successfully with `lubridate::ymd_hms()`.
6. **No missing values**: no `NA` in any column.
7. **Numeric parameter columns**: all parameter columns contain numeric values.

### Example: triggering an error

Adding an unrecognised column causes `checkASRcont()` to stop immediately:

```{r}
#| error: true

# import the data for the example
contdat_raw <- suppressWarnings(
  readxl::read_excel(contpth, na = c("NA", "na", ""), guess_max = Inf)
) |>
  mutate(across(
    where(~ inherits(.x, "POSIXct") | inherits(.x, "Date")),
    as.character
  ))

# add a column that is not in paramsASR
contdat_raw$BadColumn <- 1

checkASRcont(contdat_raw)
```

### Output format

After passing all checks, `readASRcont()` returns a data frame with:

- `Site`: site identifier (character)
- `DateTime`: combined and time-zone-aware `POSIXct` column
- One numeric column per parameter present in the input file

```{r}
head(contdat)
```

## QC threshold metadata

The metadata file includes various information for the quality control checks applied to each parameter (see the [quality control vignette](qcoverview.html) for details). Use `readASRmeta()` to import the QC threshold metadata. The function reads the workbook, runs checks via `checkASRmeta()`, and returns a formatted data frame.

```{r}
metadat <- readASRmeta(metapth)
```

### Format requirements

The workbook must contain exactly the following columns (all required; thresholds you do not want to apply should be left blank / `NA`):

| Column | Description |
|--------|-------------|
| `Site` | Site identifier matching those in the continuous data |
| `Parameter` | Parameter name matching `paramsASR$Parameter` |
| `Depth` | Sensor deployment depth (numeric; can be `NA`) |
| `GrMinFail` | Gross range, lower fail threshold |
| `GrMaxFail` | Gross range, upper fail threshold |
| `GrMinSuspect` | Gross range, lower suspect threshold |
| `GrMaxSuspect` | Gross range, upper suspect threshold |
| `SpikeFail` | Spike, absolute step size for a fail flag |
| `SpikeSuspect` | Spike, absolute step size for a suspect flag |
| `FlatFailN` | Flatline, consecutive identical observations for a fail flag |
| `FlatFailDelta` | Flatline, maximum within-run absolute change treated as "identical" for fail |
| `FlatSuspectN` | Flatline, consecutive identical observations for a suspect flag |
| `FlatSuspectDelta` | Flatline, maximum within-run absolute change treated as "identical" for suspect |
| `RoCN` | Rate of change, multiplier applied to the rolling SD (flag if `\|diff\| > SD × RoCN`) |
| `RoCHours` | Rate of change, look-back window length in hours |

### Checks performed

The `readASRmeta()` function imports the metadata and runs a series of checks using the `checkASRmeta()` function.  The checks evaluate the following and stops with an informative error if any check fails:

1. **Column names**: all columns are in the required list above.
2. **All columns present**: every required column exists.
3. **At least one parameter**: at least one value in `Parameter` matches `paramsASR$Parameter`.
4. **Parameter format**: all `Parameter` values match those in `paramsASR$Parameter`.
5. **Numeric columns**: all columns except `Site` and `Parameter` contain only numeric or missing values.

### Example: triggering an error

Supplying an unrecognised parameter name fails the parameter format check:

```{r}
#| error: true
# import the data for the example
metadat_raw <- suppressWarnings(
  readxl::read_excel(metapth, na = c("NA", "na", ""), guess_max = Inf)
)

# introduce a typo in the Parameter column
metadat_raw$Parameter[1] <- "WaterTemp"

checkASRmeta(metadat_raw)
```

### Output format

After passing all checks, `readASRmeta()` returns a data frame with the columns listed in the format requirements table above, with all threshold columns coerced to numeric.

```{r}
head(metadat)
```

The remaining functions in AquaSensR can now be used after the continuous data and metadata files are successfully imported.