---
title: "Quality control overview"
vignette: >
  %\VignetteIndexEntry{qcoverview}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false
library(AquaSensR)
```

The quality control (QC) functions in AquaSensR can be used once the required data are successfully imported into R (see the [inputs vignette](inputs.html) for details). This vignette covers the primary functions of the QC workflow:

- `utilASRflag()`: Applies four independent QC checks to a selected parameter and returns a data frame of flag results.
- `anlzASRflag()`: Produces an interactive time-series plot of those flags for visual review.

## Load the data

The examples throughout this vignette use the example files bundled with the package. Import both files before proceeding:

```{r}
contpth <- system.file("extdata/ExampleCont1.xlsx", package = "AquaSensR")
dqopth <- system.file("extdata/ExampleDQO.xlsx", package = "AquaSensR")

contdat <- readASRcont(contpth)
dqodat <- readASRdqo(dqopth)
```

`contdat` is a data frame with columns `DateTime`, and one numeric column per parameter. `dqodat` contains the parameter-specific data quality objectives (DQOs) for each check. See the [inputs vignette](inputs.html) for more information on the required formats.

## `utilASRflag()` to flag continuous data

`utilASRflag()` is the primary QC function. It applies four independent checks to the chosen parameter in `contdat`.

### Arguments

Three arguments are required for the function: 

| Argument | Description |
|----------|-------------|
| `cont` | `contdat` data frame returned by `readASRcont()` |
| `dqo` | `dqodat` data frame returned by `readASRdqo()` |
| `param` | Name of the parameter column to evaluate (must match a column in `contdat` and a `Parameter` entry in `dqodat`) |

### Basic usage

Pass the two data frames and the name of the parameter to evaluate:

```{r}
flagdat <- utilASRflag(contdat, dqodat, param = "Water_Temp_C")
head(flagdat)
```

### Output

`utilASRflag()` returns a data frame with the following columns:

| Column | Description |
|--------|-------------|
| `DateTime` | Observation timestamp |
| *`param`* | The evaluated parameter values |
| `gross_flag` | Flag from the gross range check |
| `spike_flag` | Flag from the spike check |
| `roc_flag` | Flag from the rate-of-change check |
| `flat_flag` | Flag from the flatline check |

Each flag column contains one of three values: `"pass"`, `"suspect"`, or `"fail"`. Checks are independent of each other such that a single observation can receive any combination of flags across the four columns.

If no row in `dqodat` matches a parameter a warning is returned and the function leaves all flags as `"pass"` and continues.

## QC checks explained

AquaSensR implements four QC checks that reflect widely used sensor data quality standards. The underlying concepts and code borrow heavily from the [ContDataQC](https://leppott.github.io/ContDataQC) package. All threshold values are set in the data quality objectives file and can be customised per parameter.  Manual update of these thresholds is likely necessary to avoide false positives and negatives.  Importantly, these flags require manual verification and should not be used to automatically exclude data without review.

### 1. Gross range

**DQO columns:** `GrMin`, `GrMax` (thresholds differ by row: `Flag = "Fail"` vs `Flag = "Suspect"`)

**Flag column:** `gross_flag`

The gross range check tests whether each observation falls within absolute physical or sensor limits. It is the broadest of the four checks and is intended to catch values that are simply impossible or outside the expected operating range of the instrument.

Each observation is compared to the thresholds in the two data quality objectives rows for that parameter:

- Values below `GrMin` or above `GrMax` in the `"Fail"` row return `"fail"`
- Values below `GrMin` or above `GrMax` in the `"Suspect"` row but within the fail bounds return `"suspect"`

The fail thresholds define hard physical limits (e.g., water temperature cannot be below −5 °C for a freshwater deployment). The suspect bounds are set somewhat more conservatively to flag readings that are unusual but not impossible.

Any threshold can be set to `NA` in the data quality objectives file to skip that particular flag.

Quickly view how many flags of each type were generated by the gross range check:

```{r}
# Check which observations received a gross range flag
table(flagdat$gross_flag)
```

### 2. Spike

**DQO columns:** `Spike` (threshold differs by row: `Flag = "Fail"` vs `Flag = "Suspect"`)

**Flag column:** `spike_flag`

The spike check detects sudden, anomalous jumps (either up or down) between consecutive observations. It computes the absolute difference between each reading and the one immediately before it, then compares that difference to the thresholds in the two data quality objectives rows for that parameter:

- |diff| ≥ `Spike` in the `"Suspect"` row returns `"suspect"`
- |diff| ≥ `Spike` in the `"Fail"` row returns `"fail"`

The first observation in each series has no predecessor and is always left as `"pass"`. Because the spike check flags the observation at the large step, a single anomalous reading embedded in otherwise stable data will generate two flagged observations — one for the step up (or down) to the outlier, and one for the step back to baseline.

The spike thresholds are absolute.  For example, a 5 °C step is flagged regardless of whether the surrounding series is calm or noisy. The rate-of-change check (below) evaluates potentially spurious changes when relative variability matters.

Quickly view how many flags of each type were generated by the spike check:

```{r}
table(flagdat$spike_flag)
```

### 3. Rate of change

**DQO columns:** `RoCStDv`, `RoCHours` (from the `Flag = "Suspect"` row only)

**Flag column:** `roc_flag`

The rate-of-change (RoC) check is an adaptive counterpart to the spike check. Rather than comparing against a fixed step size, the check determines whether a step is large relative to the recent variability in the series.

For each observation the function:

1. Collects all values within a trailing `RoCHours`-hour window ending at that timestamp.
2. Computes the standard deviation (SD) of those values.
3. Multiplies the SD by `RoCStDv` to produce a contextual threshold.
4. Flags the observation `"suspect"` if the absolute lag-1 difference exceeds that threshold.

At least two values must fall within the window before a standard deviation can be computed; observations with fewer window values are not flagged. Also, this check only produces `"suspect"` flags, i.e., there is no `"fail"` flag.

The key advantage over the spike check is sensitivity scaling.  During a "calm" period, a small absolute change can exceed the threshold, while during a naturally variable period (e.g., diurnal temperature swings) the threshold rises accordingly.

Quickly view how many flags of each type were generated by the rate of change check:

```{r}
table(flagdat$roc_flag)
```

### 4. Flatline

**DQO columns:** `FlatN`, `FlatDelta` (thresholds differ by row: `Flag = "Fail"` vs `Flag = "Suspect"`)

**Flag column:** `flat_flag`

The flatline check identifies periods where a sensor appears to be stuck at a constant value, which can occur from sensor fouling, burial, or loss of power. The check counts the length of "runs" of near-identical consecutive values and flags observations whose run length reaches a specified count.

A run is defined by a minimum length or count (`FlatN`) and tolerance (`FlatDelta`), each read from the appropriate data quality objectives row. The tolerance is used to identify consecutive observations considered to be "identical" if the absolute step from the previous observation is within that tolerance. This allows for a small amount of measurement noise without stopping the run count.

- Run length ≥ `FlatN` (using `FlatDelta` tolerance) from the `"Suspect"` row returns `"suspect"`
- Run length ≥ `FlatN` (using `FlatDelta` tolerance) from the `"Fail"` row returns `"fail"`

The suspect and fail thresholds are evaluated independently using their respective delta tolerances, so the two run lengths may differ. Either row can have `NA` values to skip that level.

Quickly view how many flags of each type were generated by the flatline check:

```{r}
table(flagdat$flat_flag)
```

## `anlzASRflag()` to visualise flag results

The flags generated by `utilASRflag()` can be viewed using the `anlzASRflag()` function.  This produces an interactive time-series plot:

```{r}
#| out-width: 100%
anlzASRflag(flagdat)
```

The plot shows all observations as a continuous line. Non-passing observations are overlaid as coloured markers, with colour encoding the check type and shape encoding the severity:

| Check | Colour |
|-------|--------|
| Gross range | Red |
| Spike | Orange |
| Rate of change | Purple |
| Flatline | Blue |

| Severity | Marker shape |
|----------|-------------|
| Suspect | Upward triangle |
| Fail | Cross (×) |

An observation flagged by more than one check appears as overlying markers for each check, so that all potential issues remain visible. Hovering over a marker reveals the check name, severity, parameter value, and timestamp. Items in the legend can be clicked to toggle visibility of a check or severity level, which is useful for reviewing specific flags in a busy plot.  The plot can also be zoomed and panned to focus on specific periods.